|

### 2023-06-15 manuscript.qmd
revised text

---
title: |
  Longitudinal Analysis Manuscript: Working Draft
subtitle: Longitudinal Analysis using data from the ABCD Study
short-title: Longitudinal Analysis
published: "Under review at *xxxxx Journal*"
code-repo: "Access the code, data, and analysis at <http://swhawes.github.io/test/>"
number-sections: true
number-depth: 3
correspondence-prefix: "Correspondence concerning this article should be addressed to"
author:
  - name: Samuel Hawes
    email: shawes@fiu.edu
    title: Associate Research Professor
    affiliations: 
      #- id: ds9
        name: Florida International University
        department: Center for Children & Families
        address: 11200 SW 8th St, AHC-IV
        city: Miami
        region: FL
        country: USA
        postal-code: 33199
    attributes:
      corresponding: true
  - name: Andrew K. Littlefield
    email: andrew.littlefield@ttu.edu
    affiliations: 
      #- id: ds9
        name: Texas Tech University
        department: Psychological Sciences
        city: Lubbock
        region: TX
        country: USA
  - name: Daniel A. Lopez
    email: Daniel_Lopez2@URMC.Rochester.edu
    affiliations:
      #- id: terok
        name: University of Rochester Medical Center
        department: Department of Public Health
        city: Rochester
        region: NY
        country: USA
      #- ref: ds9
  - name: Kenneth J. Sher
    email: SherK@missouri.edu
    affiliations: 
      #- id: ds9
        name: University of Missouri
        department: Psychological Sciences
        city: Columbia
        region: MO
        country: USA
  - name: Erin L. Thompson
    email: erthomps@fiu.edu
    affiliations:
      #- id: terok
        name: Florida International University
        department: Center for Children and Families
        city: Miami
        region: FL
        country: USA
  - name: Raul Gonzalez
    email: gonzara@fiu.edu
    affiliations: 
      #- id: ds9
        name: Florida International University
        department: Center for Children and Families
        city: Miami
        region: FL
        country: USA
  - name: Additional Co-authors
    email: xxxxx@fiu.edu
    affiliations:
      #- id: terok
        name: xxxxxx University
        department: xxxxxx Dept
        city: xxxx
        region: xxxxx
        country: xxxxx
  - name: Wesley K. Thompson
    email: wes.stat@gmail.com
    affiliations:
      #- id: terok
        name: Laureate Institute for Brain Research
        department: Center for Population Neuroscience and Genetics
        city: Tulsa
        region: OK
        country: USA

abstract: |
  The Adolescent Brain Cognitive Development (ABCD) Study presents a unique opportunity for researchers to investigate developmental processes in a large, diverse cohort of children and adolescents. Given the complex nature of the longitudinal data collected in the ABCD Study, researcher are likely to encounter a myriad of methodological and analytic considerations and concerns. This review provides a comprehensive examination of key issues and techniques related to longitudinal data analysis, specifically focusing on the ABCD Study. The text discusses model assumptions, common violations (e.g., independent and identically distributed residuals, *heterogeneous* *variability*) and their implications for valid inference. The importance of appropriately modeling covariance structures, understandings trade-offs between model fit and parsimony, and challenges related to sample size, attrition, missing data are highlighted. Consideration is given to the importance of selecting appropriate statistical models to account for correlations in repeated measurements and the assumptions underlying these models. The review also differentiates between linear and non-linear models in the context of continuous and discrete data, emphasizing various distributional assumptions and the necessity of choosing appropriate models and statistical methods. By addressing these complexities, the review seeks to equip researchers with the necessary knowledge and tools to make informed decisions as they navigate effectively analyzing and interpreting data available in the ABCD Study.
# thanks: |
#   Placeholder text generated at <https://vlad-saling.github.io/star-trek-ipsum/>.
# additional-info: |
#   We have no known conflict of interest to disclose.
keywords:
  - ABCD Study
  - longitudinal analyses
  - development
format:
  #hikmah-pdf: default
  hikmah-manuscript-pdf: default
  #html:
    #page-layout: article
    #toc: true
    #toc-depth: 1
    #toc-expand: 2
    #toc-title: Table of Contents
  #   code-tools:
  #     source: true
  #     toggle: false
  #     caption: none
  #   grid:
  #     sidebar-width: 200px
  #     margin-width: 0px
  pdf:
    geometry:
      - top=30mm
      - left=20mm
  #   keep_tex: yes
    #fontfamily: libertinus
    #fontsize: 20pt
#colorlinks: true
#format-links: [html, pdf]

execute:
 echo: false

# output: 
#   rticles::arxiv_article
bibliography: "references.bib"
#biblio-style: unsrt
---

@@

\label{sec:headings} [The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largest long-term investigation of neurodevelopment and child health in the United States. Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental influences on brain development and their roles in behavioral and health outcomes in adolescents [@volkow2018]. At its heart, the study is designed to chart the course of human development across multiple, interacting domains from late childhood to early adulthood and to identify factors that lead to both positive and negative developmental outcomes. Central to achieving these goals is the ABCD Study's® commitment to an open science framework designed to facilitate access to and sharing of scientific knowledge by espousing practices that increase openness, integrity, and reproducibility of scientific research (e.g., public data releases). In this sense, the ABCD Study® is a collaboration with the larger research community, with the rich longitudinal nature of the ABCD Study dataset allowing researchers to perform a variety of analyses of both methodological and substantive interest. Together, this presents a unique opportunity to significantly advance our understanding of how a multitude of biopsychosocial processes emerge and unfold across critical periods of development. In the following text, we describe methods for longitudinal analyses of ABCD Study data that can address its fundamental scientific aims, as well as challenges inherent in a large population-based long-term study of adolescents. We begin with a brief orientation to the ABCD Study and its accompanying [publicly available dataset](https://nda.nih.gov/abcd/ "Visit the ABCD Study data repository"){#ndardata}.

@@

Participants enrolled in the ABCD Study include a large cohort of youth (n=11880) aged 9-10 years at baseline and their parents/guardians. The study sample was recruited from household populations in defined catchment areas for each of the 21 study sites across the United States (information regarding funding agencies, recruitment sites, investigators, and project organization can be obtained at the ABCD Study website). The ABCD Study is collecting longitudinal data on a rich variety of outcomes that will enable the construction of realistically-complex etiological models by incorporating factors from many domains simultaneously. Each new wave of data collection provides the building blocks for conducting probing longitudinal analyses that allow us to characterize normative development, identify variables that presage deviations from prototypic development, and assess a range of outcomes associated with variables of interest. This data includes a neurocognitive battery [@luciana2018a; @thompson2019], mental and physical health assessments [@barch2018], measures of culture and environment [@zucker2018], substance use \[add citation\], biospecimens [@uban2018], structural and functional brain imaging [@casey2018; @hagler2019], geolocation-based environmental exposure data, wearables, and mobile technology [@bagot2018], and whole genome genotyping [@loughnan2020]. Many of these measures are collected at in-person annual visits, with brain imaging collected at baseline and every other year going forward. A limited number of assessments are collected in semi-annual telephone interviews between in-person visits. Data are publicly released on an annual basis through the NIMH Data Archive. By necessity, the study's earliest data releases were cross-sectional (i.e., the baseline data), however, the most recent public data release (NDA Release 4.0) contains data collected across three annual assessments, including two imaging assessments (baseline and year 2 follow-up visits).

@@

This manuscript will go through the following sections :xxxxxx, xxxx, xxxxx.
# Part I: Longitudinal Research
# Part II: Longitudinal Data
# Part III: Longitudinal Analysis
Overview of **concepts and considerations specific to developmentally-focused longitudinal research**

@@

Aims/Goals: Use this xxxxx during any xxxxx and you're virtually guaranteed to xxxxx

-   Concepts and Considerations when working with large and multimodal longitudinal data

-   Concepts and considerations of developmental research (longitudinal research via developmental perspective)

-   Learn specific methods

-   Use our open-source resources to get started more quickly

-   Learn How to Apply xxxxx to large longitudinal data such as the ABCD Study

identify common xxxxx of longitianl research

1.3 Objectives and scope
The focus of this xxxx is on the analysis of event history data using
xxxx model with the objectives
• to demonstrate the use of event history analysis in the analysis
of xxxxx, xxxx and to identify the methodological problems in the analysis,
• to propose procedures to choose a basic time scale,
• to discuss the connections between the methods for modeling
dual time scales and to perform quantitative comparisons between them,
• to compare existing methods to deal with longitudinal measurements in the xxxx model with two proposed methods.

:::

# Part I: Longitudinal Research

@@

There are several important concepts to consider when conducting longitudinal analyses in a developmental context. These include different ways of thinking about developmental course, whether certain periods of development are relatively sensitive or insensitive to various types of insults or stressors, whether some time periods or situations inhibit the expression of individual differences due to extreme environmental pressures, and whether the same behavior manifested at different times represent the same phenomenon or different ones. Further, in the case of developmentally focused longitudinal research, each new measurement occasion not only provides a more extended portrait of the child's life course (and not just characterize growth during this period but also assesses the durability/chronicity of prior effects/consequences) but also brings with it greater methodological opportunities to exploit the statistical properties of longitudinal data in the furtherance of critical scientific questions. That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows, assuming we have assessed the "right" variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest. Appreciation of these and other issues can help to guide the analysis and interpretation of data and aid translation to clinical and public health applications.

**Vulnerable periods**. Development normatively progresses from less mature to more mature levels of functioning. However, unique epochs and experiences can alter the course of this idealized form of development. Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development \[add citation\]; or, similarly, experimental research on rodents that shows rodent brains to be especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development (corresponding to early adolescence in humans)\[add citation\]. These examples highlight the importance of considering the role of vulnerable periods -- temporal windows of rapid brain development or remodeling during which the effects of environmental stimuli (e.g. cannabis exposure) on the developing brain may be particularly pronounced-- when trying to establish an accurate understanding of the association between exposures and outcomes.

**Developmental disturbances**. Whereas vulnerable periods heighten neurobiological susceptibility to environmental influences, at other times environmental pressures will tend to suppress stability and disrupt the orderly stochastic process of normative development (e.g., xxx-xxx). This situation reflects a developmental disturbance in that the normal course of development is "disturbed" for some time by some time-limited process. In such cases, we might find that prediction of behavior in the period of the disturbance is reduced and/or, similarly, the behavior exhibited during the disturbance might have less predictive power with respect to distal outcomes compared to the behavior exhibited before and following the disrupted period. That is, once the environmental stimuli are removed (or the individual is removed from the environment), individual differences are again more freely expressed and the autoregressive effects increase to levels similar to those before entering the environment.

**Developmental snares and cascade effects**. Normative development can also be upended by experiences (e.g., drug use) that, through various mechanisms, disrupt the normal flow of development wherein each stage establishes a platform for the next. For instance, substance use could lead to association with deviant peers, precluding opportunities for learning various adaptive skills and prosocial behaviors, in effect, creating a "snare" that retards psychosocial development. Relatedly, the consequences of these types of events can cascade (e.g., school dropout, involvement in the criminal justice system) so that the effects of the snare are amplified. Although conceptually distinct from vulnerable periods, both of these types of developmental considerations highlight the importance of viewing behavior in the context of development and the importance of attempting to determine how various developmental pathways unfold.

**Distinguishing developmental change from experience effects**. One can often observe systematic changes over time in a variable of interest and assume this change is attributable to development. To this point, cognitive abilities (e.g, verbal ability, problem-solving) normatively grow earlier in development and often decline in late life (e.g., memory, speed of processing). However, the observed patterns of growth and decline often differ between cross-sectional vs. longitudinal effects [@salthouse2014] where subjects gain increasing experience with the assessment with each successive measurement occasion. Such experience effects on cognitive functioning have been demonstrated in adolescent longitudinal samples similar to ABCD [@sullivan2017] and highlight the need to consider these effects and address them analytically. In the case of performance-based measures [e.g., matrix reasoning related to neurocognitive functioning; see @salthouse2014], this can be due to "learning" the task from previous test administrations (e.g., someone taking the test a second time performs better than they did the first time simply as a function of having taken it before). Even in the case of non-performance-based measures (e.g., levels of depression), where one cannot easily make the argument that one has acquired some task-specific skill through learning, it has been observed that respondents tend to endorse lower levels on subsequent assessments [e.g., @beck1961; see @french2010] and this phenomenon has been well documented in research on structured diagnostic interviews [@robins1985]. While it is typically assumed that individuals are rescinding or telling us less information on follow-up interviews, there is reason to suspect that in some cases the initial assessment may be artefactually elevated [see @shrout2018a]. Some designs (specifically, accelerated longitudinal designs) are especially well suited for discovering these effects and modeling them. While ABCD was not designed as an accelerated longitudinal design, the variability in age at the time of baseline recruitment (9 years, 0 months to 10 years, 11 months) allows some measures, collected every year, to be conceptualized as an accelerated longitudinal design. Moreover, it is possible that in later waves, patterns of longitudinal missing data will allow some analyses to assess the confounded effects of age and the number of prior assessments. However, ABCD is fundamentally a single-cohort, longitudinal design, where a number of prior assessments and age are highly confounded, and for, perhaps, most analyses, the possible influence of experience effects needs to be kept in mind.

# Part II: Longitudinal Data

@@

**Defining Features of Longitudinal Data Analysis.** The hallmark characteristic of longitudinal data analysis is its application to repeated assessments of the same assessment targets (e.g., individuals, families) across time. While the primary reason for collecting longitudinal data is in pursuit of addressing scientific questions, from a methodological perspective, having multiple observations over time allows researchers to identify potentially problematic observations when highly improbable longitudinal patterns are observed. That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows assuming we have assessed the "right" variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest .

["traditional approaches": marginal models]
[random effects: Growth Curves- SEM (lgcm) vs ME (lmm)]
[(between- and within-person levels?)]

### Modeling Data Across Two Time Points versus Three or More Time Points.

@@

If one were to try to sum up what development in a living organism is exactly, one could plausibly argue it's the characterization of stability and change as the organism traverses the life course. There are a few different ways to think of stability (and change). Consider we measure the height of all youth in a 6th-grade class, once in the fall at the beginning of the school year and once again in the spring at the end of the school year. A common first step may be to compare the class's average height values obtained at these two different measurement occasions. This comparison of the average scores for the same group of individuals at multiple time points is referred to as "mean-level" stability as it provides information about continuity and change in the group level of an outcome of interest (e.g., height) over time. Another type of stability involves calculating the correlation between the values obtained at different time points (e.g., 'height in the fall' with 'height in the spring'). This type of "rank-order" stability evaluates between-individual change by focusing on the degree to which individuals retain their relative placement in a group across time. Consider, someone who is the shortest person in their class in 6th grade may grow considerably over the school year (i.e., exhibit mean level change), but remain the shortest person among their classmates. That is, the individual is manifesting a type of rank-order stability. Both types of stability and change are important. Mean-level change in certain traits might help to explain why, in general, individuals are particularly vulnerable to social influences at some ages more than others; rank order change might help to quantify the extent to which certain characteristics of the individual are more trait-like. For example, in some areas of development, there is considerable mean-level change that occurs over time (e.g., changes in Big 5 personality traits), but relatively high rank-order stability. Despite the useful information afforded by examining mean-level and rank-order change, these approaches are limited in that they provide little information about patterns of "within-individual" change and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest.

There is growing recognition that statistical models commonly applied to longitudinal data often fail to comport with the developmental theory they are being used to assess (e.g., Curran, Lee, Howard, Lane, & MacCallum, 2012; Hoffman, 2015; Littlefield et al., 2021. Specifically, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person (i.e., intraindividual) processes (e.g., how phenotypes change or remain stable within individuals over time) [e.g., see @curran2011]. Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures. As a result, experts increasingly caution about the need to xxxxxxxx [add citation]. Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield2021; @orth2021]. Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual's mean (or trait) level across time [e.g., RI-CLPM, @hamaker2015; LCM-SR, @curran2014a]. It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield2021, for further discussion].

This is illustrated well by @fig-LDAstruct.

@@

Many statistical models assume certain characteristics about the data to which they are being applied. As an example, common assumptions of parametric statistical models include normality, linearity, and equality of variances. These assumptions must be carefully considered before conducting analysis so that valid inferences can be made from the data; that is, violation of a model's assumptions can substantively alter the interpretation of results. Similarly, statistical models employed in the analyses of longitudinal data often entail a range of assumptions that must be closely inspected. One central issue for repeated measurements on an individual is how to account for the correlated nature of the data; another common feature of longitudinal data is heterogeneous variability; that is, the variance of the response changes over the duration of the study. Traditional techniques, such as a standard regression or ANOVA model, assume residuals are independent and thus are inappropriate for designs that assess (for example) the same individuals across time. That is, given the residuals are no longer independent, the standard errors from the models are biased and can produce misleading inferential results. Although there are formal tests of independence for time series data (e.g., the Durbin-Watson statistic [@durbin1950]), more commonly independence is assumed to be violated in study designs with repeated assessments. Therefore, an initial question to be addressed by a researcher analyzing prospective data is how to best model the covariance structure of said data.

@@

##### SEM

@@

##### Advanced SEM

Advanced structural equation modeling (SEM) approaches, such as the Random-Intercept Cross-Lagged Panel Model (RI-CLPM) and Latent Curve Models with Structured Residuals (LCM-SR), have emerged to address more complex research questions and data structures in longitudinal analysis. These advanced models extend traditional SEM techniques, enabling researchers to disentangle within-person and between-person effects, as well as capture additional time-specific dependencies and associations that may not be accounted for by the latent growth factors.

>>>>

---
title: |
  Longitudinal Analysis Manuscript: Working Draft
subtitle: Longitudinal Analysis using data from the ABCD Study
short-title: Longitudinal Analysis
code-repo: "Access the code, data, and analysis at <http://swhawes.github.io/test/>"
number-sections: true
number-depth: 3
correspondence-prefix: "Correspondence concerning this article should be addressed to"
author:
  - name: Samuel Hawes
    email: shawes@fiu.edu
    title: Associate Research Professor
    affiliations:
      #- id: ds9
        name: Florida International University
        department: Center for Children & Families
        address: 11200 SW 8th St, AHC-IV
        city: Miami
        region: FL
        country: USA
        postal-code: 33199
    attributes:
      corresponding: true
  - name: Andrew K. Littlefield
    email: andrew.littlefield@ttu.edu
    affiliations:
      #- id: ds9
        name: Texas Tech University
        department: Psychological Sciences
        city: Lubbock
        region: TX
        country: USA
  - name: Daniel A. Lopez
    email: Daniel_Lopez2@URMC.Rochester.edu
    affiliations:
      #- id: terok
        name: University of Rochester Medical Center
        department: Department of Public Health
        city: Rochester
        region: NY
        country: USA
      #- ref: ds9
  - name: Kenneth J. Sher
    email: SherK@missouri.edu
    affiliations:
      #- id: ds9
        name: University of Missouri
        department: Psychological Sciences
        city: Columbia
        region: MO
        country: USA
  # - name: Erin L. Thompson
  #   email: erthomps@fiu.edu
  #   affiliations:
  #     #- id: terok
  #       name: Florida International University
  #       department: Center for Children and Families
  #       city: Miami
  #       region: FL
  #       country: USA
  # - name: Raul Gonzalez
  #   email: gonzara@fiu.edu
  #   affiliations:
  #     #- id: ds9
  #       name: Florida International University
  #       department: Center for Children and Families
  #       city: Miami
  #       region: FL
  #       country: USA
  - name: Additional Co-authors
    email: xxxxx@fiu.edu
    affiliations:
      #- id: terok
        name: xxxxxx University
        department: xxxxxx Dept
        city: xxxx
        region: xxxxx
        country: xxxxx
  - name: Wesley K. Thompson
    email: wes.stat@gmail.com
    affiliations:
      #- id: terok
        name: Laureate Institute for Brain Research
        department: Center for Population Neuroscience and Genetics
        city: Tulsa
        region: OK
        country: USA

format:
  docx:
    toc: false
  elsevier-pdf:
    journal:
      cite-style: number
      formatting: preprint
      model: 3p
      layout: twocolumn
      #cite-style: number
    include-in-header: 
      text: |
        \newpageafter{author}
  html:
    page-layout: article
    toc: true
    toc-depth: 1
    toc-expand: 2
    toc-title: Table of Contents
    code-tools:
      source: true
      toggle: false
      caption: none
    grid:
      sidebar-width: 200px
      margin-width: 0px

#fontfamily: libertinus
#fontsize: 20pt
#colorlinks: true
#format-links: [html, pdf]
#execute:
  #echo: false
#bibliography: "references.bib"
#biblio-style: unsrt

abstract: |
  The Adolescent Brain Cognitive Development (ABCD) Study presents a unique opportunity for researchers to investigate developmental processes in a large, diverse cohort of children and adolescents. Given the complex nature of the longitudinal data collected in the ABCD Study, researchers are likely to encounter a myriad of methodological and analytic considerations and concerns. This review provides a comprehensive examination of key issues and techniques related to longitudinal data analysis, with a specific focus on use with the ABCD Study dataset. The text discusses model assumptions, common violations (e.g., independent and identically distributed residuals, heterogeneous variability), and their implications for valid inference. The importance of appropriately modeling covariance structures, understandings trade-offs between model fit and parsimony, and challenges related to sample size, attrition, and missing data are highlighted. Consideration is given to the importance of selecting appropriate statistical models to account for correlations in repeated measurements and the assumptions underlying these models. The review also differentiates between linear and non-linear models in the context of continuous and discrete data, emphasizing various distributional assumptions and the necessity of choosing appropriate models and statistical methods. By addressing these complexities, the review seeks to equip researchers with the necessary knowledge and tools to make informed decisions as they navigate effectively analyzing and interpreting data available in the ABCD Study.

keywords:
- ABCD Study
- longitudinal analyses
- development

---

@@

\label{sec:headings} [The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largestz long-term investigation of neurodevelopment and child health in the United States. Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental influences on brain development and their roles in behavioral and health outcomes in adolescents [@volkow2018]. At its heart, the study is designed to chart the course of human development across multiple, interacting domains from late childhood to early adulthood and to identify factors that lead to both positive and negative developmental outcomes. Central to achieving these goals is the ABCD Study's® commitment to an open science framework designed to facilitate access to and sharing of scientific knowledge by espousing practices that increase openness, integrity, and reproducibility of scientific research (e.g., public data releases). In this sense, the ABCD Study® is a collaboration with the larger research community, with the rich longitudinal nature of the ABCD Study dataset allowing researchers to perform a variety of analyses of both methodological and substantive interest. Together, this presents a unique opportunity to significantly advance our understanding of how a multitude of biopsychosocial processes emerge and unfold across critical periods of development. In the following text, we describe methods for longitudinal analyses of ABCD Study data that can address its fundamental scientific aims, as well as challenges inherent in a large population-based long-term study of adolescents. We begin with a brief orientation to the ABCD Study and its accompanying [publicly available dataset](https://nda.nih.gov/abcd/ "Visit the ABCD Study data repository"){#ndardata}.

@@

Participants enrolled in the ABCD Study include a large cohort of youth (n=11,880) aged 9-10 years at baseline and their parents/guardians. The study sample was recruited from household populations in defined catchment areas for each of the 21 study sites across the United States (information regarding funding agencies, recruitment sites, investigators, and project organization can be obtained at the ABCD Study website). The ABCD Study is collecting longitudinal data on a rich variety of outcomes that will enable the construction of complex etiological models by incorporating factors from many domains simultaneously. Each new wave of data collection provides the building blocks for conducting probing longitudinal analyses that allow us to characterize normative development, identify variables that presage deviations from prototypic development, and assess a range of outcomes associated with variables of interest. This data includes a neurocognitive battery [@luciana2018a; @thompson2019], mental and physical health assessments [@barch2018], measures of culture and environment [@zucker2018], substance use [@lisdahl2021], biospecimens [@uban2018], structural and functional brain imaging [@casey2018; @hagler2019], geolocation-based environmental exposure data, wearables, and mobile technology [@bagot2018], and whole genome genotyping [@loughnan2020]. Many of these measures are collected at in-person annual visits, with brain imaging collected at baseline and every other year going forward. A limited number of assessments are collected in semi-annual telephone interviews between in-person visits. Data are publicly released on an annual basis through the NIMH Data Archive. By necessity, the study's earliest data releases were cross-sectional (i.e., the baseline data), however, the most recent public data release (NDA Release 4.0) contains data collected across three annual assessments, including two imaging assessments (baseline and year 2 follow-up visits).

@@

@@

• Part I. Longitudinal Research

  -   Identify fundamental concepts

• Part II. Longitudinal Data

  -   Highlight key challenges

• Part III. Longitudinal Analysis

  -   Learn methods & analysis to apply

• Part IV. Supplemental materials

  -   Make use of linked open-source resources
:::

# Longitudinal Research

@@

There are several important concepts to consider when conducting longitudinal analyses in a developmental context. These include different ways of thinking about developmental course, whether certain periods of development are relatively sensitive or insensitive to various types of insults or stressors, whether some time periods or situations inhibit the expression of individual differences due to extreme environmental pressures, and whether the same behavior manifested at different times represent the same phenomenon or different ones. Further, in the case of developmentally focused longitudinal research, each new measurement occasion not only provides a more extended portrait of the child’s life course but also brings with it greater methodological opportunities to make use of statistical models that distinguish within and between subject effects and that loosen constraints that could be imposed on the furtherance of critical scientific questions. Appreciation of these and other issues can help to guide the analysis and interpretation of data and aid translation to clinical and public health applications.

### Vulnerable periods.

Development normatively progresses from less mature to more mature levels of functioning. However, unique epochs and experiences can alter the course of this idealized form of development. Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development [@arseneault2002; @bechtold2016; @hasan2020; @semple2005]; or, similarly, experimental research on rodents that shows rodent brains to be especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development (corresponding to early adolescence in humans) [@spear2016; @crews2000; @ji2018]. These examples highlight the importance of considering the role of vulnerable periods -- temporal windows of rapid brain development or remodeling during which the effects of environmental stimuli (e.g. cannabis exposure) on the developing brain may be particularly pronounced-- when trying to establish an accurate understanding of the association between exposures and outcomes.

### Developmental disturbances.

Whereas vulnerable periods heighten neurobiological susceptibility to environmental influences, at other times environmental pressures will tend to suppress stability and disrupt the orderly stochastic process of normative development [e.g., @schulenberg2019]. This situation reflects a developmental disturbance in that the normal course of development is "disturbed" for some time by some time-limited process. In such cases, we might find that prediction of behavior in the period of the disturbance is reduced and/or, similarly, the behavior exhibited during the disturbance might have less predictive power with respect to distal outcomes compared to the behavior exhibited before and following the disrupted period. That is, once the environmental stimuli are removed (or the individual is removed from the environment), patterns of individual differences (and autoregressive effects) increase to levels similar to those prior to entering the environment.

### Developmental snares and cascade effects.

Normative development can also be upended by experiences (e.g., drug use) that, through various mechanisms, disrupt the normal flow of development wherein each stage establishes a platform for the next. For instance, substance use could lead to association with deviant peers, precluding opportunities for learning various adaptive skills and prosocial behaviors, in effect, creating a "snare" that retards psychosocial development such as maturing out of adolescent antisocial behavior [@moffitt2015]. Relatedly, the consequences of these types of events can cascade (e.g., school dropout, involvement in the criminal justice system) so that the effects of the snare are amplified [e.g., @masten2005; @rogosch2010]. Although conceptually distinct from vulnerable periods, both of these types of developmental considerations highlight the importance of viewing behavior in the context of development and the importance of attempting to determine how various developmental pathways unfold.

### Distinguishing developmental change from experience effects.

One can often observe systematic changes over time in a variable of interest and assume this change is attributable to development. To this point, cognitive abilities (e.g, verbal ability, problem-solving) normatively grow earlier in development and often decline in late life (e.g., memory, speed of processing). However, the observed patterns of growth and decline often differ between cross-sectional vs. longitudinal effects [@salthouse2014] where subjects gain increasing experience with the assessment with each successive measurement occasion. Such experience effects on cognitive functioning have been demonstrated in adolescent longitudinal samples similar to ABCD [@sullivan2017] and highlight the need to consider these effects and address them analytically. In the case of performance-based measures [e.g., matrix reasoning related to neurocognitive functioning; see @salthouse2014], this can be due to "learning" the task from previous test administrations (e.g., someone taking the test a second time performs better than they did the first time simply as a function of having taken it before). Even in the case of non-performance-based measures (e.g., levels of depression), where one cannot easily make the argument that one has acquired some task-specific skill through learning, it has been observed that respondents tend to endorse lower levels on subsequent assessments [e.g., @beck1961; see @french2010] and this phenomenon has been well documented in research on structured diagnostic interviews [@robins1985]. While it is typically assumed that individuals are rescinding or telling us less information on follow-up interviews, there is reason to suspect that in some cases the initial assessment may be artefactually elevated [see @shrout2018a]. Some designs (specifically, accelerated longitudinal designs) are especially well suited for discovering these effects and modeling them. While ABCD was not designed as an accelerated longitudinal design, the variability in age at the time of baseline recruitment (9 years, 0 months to 10 years, 11 months) allows some measures, collected every year, to be conceptualized as an accelerated longitudinal design. It is possible that in later waves, analyses will allow for disaggregating the confounded effects of age and the number of prior assessments. However, ABCD is fundamentally a single-cohort, longitudinal design, where a number of prior assessments and age are highly confounded, and for, perhaps, most analyses, the possible influence of experience effects needs to be kept in mind.

# Longitudinal Data

@@

The hallmark characteristic of longitudinal data analysis is its application to repeated measurements of the same assessment targets (e.g., individuals, families) across time. While the primary reason for collecting longitudinal data is in pursuit of addressing scientific questions, from a methodological perspective, having multiple observations over time allows researchers to identify potentially problematic observations when highly improbable longitudinal patterns are observed. That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows assuming we have assessed the "right" variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest.

### Two Time Points versus Three or More.

@@

If one were to try to sum up what development in a living organism is exactly, one could plausibly argue it's the characterization of stability and change as the organism traverses the life course. There are a few different ways to think of stability (and change). Consider we measure the height of all youth in a 6th-grade class, once in the fall at the beginning of the school year and once again in the spring at the end of the school year. A common first step may be to compare the class's average height values obtained at these two different measurement occasions. This comparison of the average scores for the same group of individuals at multiple time points is referred to as "mean-level" stability as it provides information about continuity and change in the group level of an outcome of interest (e.g., height) over time. Another type of stability involves calculating the correlation between the values obtained at different time points (e.g., 'height in the fall' with 'height in the spring'). This type of "rank-order" stability evaluates between-individual change by focusing on the degree to which individuals retain their relative placement in a group across time. Consider someone who is the shortest person in their class in 6th grade may grow considerably over the school year (i.e., exhibit mean level change), but remain the shortest person among their classmates. That is, the individual is manifesting a type of rank-order stability. Both types of stability and change are important. Mean-level change in certain traits might help to explain why, in general, individuals are particularly vulnerable to social influences  at some ages more than others; rank order change might help to quantify the extent to which certain characteristics of the individual are more trait-like. For example, in some areas of development, there is considerable mean-level change that occurs over time (e.g., changes in Big 5 personality traits), but relatively high rank-order stability. Despite the useful information afforded by examining mean-level and rank-order change, these approaches are limited in that they provide little information about patterns of "within-individual" change and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest[@curran2011].

![Types of Stability and Change](./figures/StabilityChange.png){#fig-StablityChange fig-alt="Stability & Change Diagram." width="300" height="300" fig-align="left"}

There is growing recognition that statistical models commonly applied to longitudinal data often fail to align with the developmental theory they are being used to assess [e.g., @curran2012; @hoffman2015; @littlefield2021]. Specifically, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person (i.e., intraindividual) processes (e.g., how phenotypes change or remain stable within individuals over time) [e.g., see @curran2011]. Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures. Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield2021; @orth2021]. Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual's mean (or trait) level across time [e.g., RI-CLPM, @hamaker2015; LCM-SR, @curran2014a]. It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield2021, for further discussion].

@@

Many statistical models assume certain characteristics about the data to which they are being applied. Common assumptions of parametric statistical models include normality, linearity, and equality of variances. These assumptions must be carefully considered before conducting analysis so that valid inferences can be made from the data; that is, violation of a model's assumptions can substantively alter the interpretation of results. For example, longitudinal data can exhibit heterogeneous variability (i.e., the variance of the response changes over the duration of the study) that will need to be accounted for. Similarly, statistical models used to analyze longitudinal data often entail a range of assumptions that must be closely inspected. One common prerequisite for analyses of repeated measures data is the assessment of longitudinal measurement invariance, which helps to establish whether or not a construct is measured consistently over time [@liu2017; @van2015; @willoughby2012]. Establishing longitudinal measurement invariance, in turn, provides researchers with greater confidence that any change over time identified for a construct is attributable to individual change rather than a measurement artifact. Another central issue for repeated measurements on an individual is how to account for the correlated nature of the data. Traditional techniques, such as a standard regression or ANOVA model, assume residuals are independent and thus are inappropriate for designs that assess (for example) the same individuals across time. That is, given the residuals are no longer independent, the standard errors from the models are biased and can produce misleading inferential results. Although there are formal tests of independence for time series data (e.g., the Durbin-Watson statistic [@durbin1950]), more commonly independence is assumed to be violated in study designs with repeated assessments. Therefore, an initial question to be addressed by a researcher analyzing prospective data is how to best model the covariance structure of said data.

@@

#### SEM

@@

#### Advanced SEM

Advanced structural equation modeling (SEM) approaches, such as the random-intercept cross-lagged panel model (RI-CLPM) and latent curve models with structured residuals (LCM-SR), have emerged to address more complex research questions and data structures in longitudinal analysis. These advanced models extend traditional SEM techniques, enabling researchers to disentangle within-person and between-person effects, as well as capture additional time-specific dependencies and associations that may not be accounted for by the latent growth factors.
